---
title: "TITLE HERE"
author: "YOUR NAME"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: pdf
---
```{r setup}
#| include: false
# Don't touch this chunk. This is setup code to hide your code chunks automatically (without needing to specify options on each chunk)
knitr::opts_chunk$set(
  # do not show code chunks
  echo = FALSE,
  # do not print warnings / console messages from code
  warning = FALSE,
  message = FALSE,
  # better looking default plot dimensions
  fig.width = 8,
  fig.height = 5,
  dpi = 200
)
library(fixest)
library(ggplot2)
```

```{r}
# Load data
```





<!-- DELETE THIS AND BELOW -->

*This is an example section to give you an idea of how your results should look. Make sure not to include this in your final report.*

## How does a car's mpg relate to its horespower?

I am analyzing a dataset consisting of different models of cars and their characteristics. 
I wanted to see how a car's horsepower and their mileage relate to one other. 

```{r}
corr_mpg_hp <- cor(x = mtcars$hp, y = mtcars$mpg)
```
```{r}
est <- feols(mpg ~ hp, data = mtcars)
marginal_effect_of_50hp <- round(abs(coef(est)["hp"] * 50), 1)
```

In @fig-hp-vs-mpg, I am plotting the car's horespower and their miles per gallon. 
You can see a strong negative correlation between the two meaning that cars with higher horespower have lower miles per gallon.
This correlation is strong with an estimated correlation of `r round(corr_mpg_hp, 2)`.
A simple regression analysis suggests that a car with an extra 50 horsepower on average is predicted to have `r marginal_effect_of_50hp` lower miles per gallon.


```{r}
#| echo: false
#| fig.cap: "Relationship between Horsepower and Miles per gallon"
#| label: fig-hp-vs-mpg
ggplot() +
  geom_point(
    aes(x = hp, y = mpg),
    data = mtcars
  ) +
  labs(
    x = "Car's horsepower",
    y = "Car's miles per gallon",
    title = "Cars with higher horsepower have lower mileage"
  ) +
  theme_light(base_size = 14)
```




