---
title: "Regression in Practice"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  pdf:
    fig-width: 8
    fig-height: 5
---

# Regression in Practice

This R day is going to have you apply different regression strategies presented in topic 04. 
An emphasis will be placed on interpreting and presenting regression results. 
For this assignment, I have assembled a dataset consisting of homes in Massachusetts. 

There are a lot of variables in this dataset, so please see the separate codebook for details.

Because this data is large (almost 2 million rows), I am using the `parquet` file format. 
You can load this with the `read_parquet` function from the `arrow` package. 
You might have to change your working directory in quarto to the folder where this `.qmd` file is located.

```{r setup}
#| message: false
library(ggplot2)
library(fixest)
library(arrow)

df <- read_parquet("data/MA_parcels_sample.parquet")
```

## Basic exploration of variables

One of the main variables we will use in this analysis is the assessed value of the home (`total_value`) and the size of the land (`lot_size_acres`).
Like good data analysts, let's create histograms of both variables. 
Make sure to include high-quality axis labels:
Comment on the distribution.

```{r}

```

*Answer:*




Now, let's look at a few explanatory variables. 
Use the `table` function to explore the distribution of `n_rooms` and make a histogram of `z_mls_acres`.

```{r}

```

*Answer:*




## Relationship between lot size and building value

We are going to explore the relationship between the lot size and the value of the building. 
We will make a scatter plot to start our analysis.
However, there is way too many points to plot in the dataset, so we will sample a random 25000 rows.

```{r}
set.seed(20260312)
sample_df <- df[sample(1:nrow(df), 25000), ]
```


Make this plot:

```{r}
ggplot() +
  geom_point(aes(x = lot_size_acres, y = total_value), data = df) +
  geom_smooth(
    aes(x = lot_size_acres, y = total_value),
    data = df,
    formula = y ~ x
  )

```





### Bivariate regression

First, regress building value on the lot size linearly using `feols` from the `fixest` package. 
Interpret the regression coefficient and comment on it's statistical significance.
Make sure to use `vcov = "HC1"` to get robust standard errors.

```{r}
# Linear regression of building value on lot size
model1 <- feols(total_value ~ lot_size_acres, data = df, vcov = "HC1")
summary(model1)
```

*Answer:*




### Multivariate regression

The size of the lot is going to be correlated with how close to downtown Boston it is. 
Let's control for `dist_to_cbd_km` and compare results.
Run this regression in `model2` and then use `etable` to compare estimates.
Compare the two coefficients. 
Why do you think the coefficient on lot size grew after adding this control? 

```{r}

```

*Answer:*


### Binned version

Now, let's use `i(dist_to_cbd_bin)` instead of `dist_to_cbd_km` to control for "bins" of cbd distance. Do the results look similar?

```{r}

```

*Answer:*


### Forecasting

Now, let's try and show what our forecasted values will be. 
To do so, let's create a `data.frame` containing a grid of values of `lot_size_acres`. 
We can do that with `seq(0, 1.5, 0.01)`.
Call this `pred_grid`.
Create forecasts from your simple linear model using `predict(..., newdata = pred_grid)`.
Store the forecasts back into the `pred_grid` data frame.

To create a plot, first copy your `plot` call above to create the sample scatterplot.
Then, use the `lines` function to overlay your forecasted line of best fit. 
Give this line a distinct color and make the line thicker by adding the arguments `col = "red", lwd = 2`.

```{r}
# Create prediction grid
pred_grid <- data.frame(
  lot_size_acres = seq(0, 1.5, 0.01)
)

# Generate predictions
pred_grid$predicted_value <- predict(model1, newdata = pred_grid)

# Plot scatter with regression line
plot(
  sample_df$lot_size_acres,
  sample_df$total_value,
  main = "Scatterplot of Total Value vs Lot Size",
  xlab = "Lot Size (Acres)",
  ylab = "Total Value",
  pch = 19,
  col = rgb(0, 0, 0, 0.1)
)
lines(pred_grid$lot_size_acres, pred_grid$predicted_value, col = "red", lwd = 2)
```

*Answer:*




### Forecasting with confidence intervals

We want to discuss uncertainty around our estimates.
Add the option `se.fit = TRUE` to your `predict` call to get a standard error for the predicted value. 
Present your result with a 95% confidence interval.

```{r}
(preds <- predict(
  model1,
  newdata = pred_grid,
  se.fit = TRUE
))
pred_grid$predicted_value <- preds$fit
pred_grid$se <- preds$se.fit

## Calculate 95% CI
# pred_grid$ci_lower <-
# pred_grid$ci_upper <-

## Plot
ggplot() +
  geom_point(
    aes(x = lot_size_acres, y = total_value),
    data = df,
    shape = 19,
    alpha = 0.1
  ) +
  geom_ribbon(
    aes(x = lot_size_acres, ymin = ci_lower, ymax = ci_upper),
    data = pred_grid,
    color = "blue"
  ) +
  geom_line(
    aes(x = lot_size_acres, y = predicted_value),
    data = pred_grid,
  )
```

*Answer:*



### Polynomial

Estimate a new model with a 4th-order polynomial of `lot_size_acres`. 
You can do this simply with the `poly` function.
Forecast this model using `pred_grid` and create a new forecast plot similar to above. 
Add lines for both the linear fit (in red) and the polynomial fit (in blue)

```{r}
## Run 4th-order polynomial regression

## Generate polynomial predictions
# pred_grid$predicted_poly <-

## Plot both linear and polynomial fits
ggplot() +
  geom_point(
    aes(x = lot_size_acres, y = total_value),
    data = df,
    shape = 19,
    alpha = 0.1
  ) +
  geom_line(
    aes(x = lot_size_acres, y = predicted_value, color = "Linear"),
    data = pred_grid
  ) +
  geom_line(
    aes(x = lot_size_acres, y = predicted_poly, color = "Polynomial"),
    data = pred_grid
  )
```

*Answer:*


### Adding other predictors

Now, let's try to add the number of rooms and the building area to our polynomial model.
Let's create indicatro variables for each value of `n_rooms` using the `i` function.
Control for building_area linearly.

```{r}
# Model with indicator variables for n_rooms and linear building_area
model_final <- feols(
  total_value ~ poly(lot_size_acres, 4) + i(n_rooms) + building_area,
  data = df,
  vcov = "HC1"
)
summary(model_final)
```

#### Exercise

Answer the following questions:

1. What is the omitted category for `n_rooms`? 

2. Interpret in words the coefficient on `n_rooms::7` and comment on it's statistical significance.

3. We could have run a regression controlling linearly for the number of rooms (treating it as a continuous measure). Do your estimates for `i(n_rooms)` make you comfortable with making that simplifying assumption?

4. What is the estimated return on square footage for homes in Massachusetts? To help put this result in context, multiply this coefficient by the standard deviation of `building_area`. Your interpretation will be a for "1 standard deviation increase in `building_area`"

*Answer:*



### Interactions

We've seen how to include indicator variables for discrete variables like `n_rooms`. Now let's explore how the effect of one variable might depend on the level of another variable through interaction terms.

#### Interaction of two discrete variables

Let's examine whether the effect of having 7 rooms (`n_rooms::7`) on home value differs depending on whether 2-family homes are permitted by right (`z_permit_2fam`). We'll create an interaction between these two discrete variables.

```{r}
# Model with interaction between n_rooms and z_permit_2fam
model_interaction1 <- feols(
  total_value ~ poly(lot_size_acres, 4) +
    i(n_rooms) +
    i(z_permit_2fam) +
    i(n_rooms, z_permit_2fam) +
    building_area,
  data = df,
  vcov = "HC1"
)
summary(model_interaction1)
```

**Questions to consider:**
1. How do you interpret the coefficient on the interaction term `n_rooms::7:z_permit_2fam::By Right`?
2. Is this interaction statistically significant?
3. What does this tell us about how the value of additional rooms depends on zoning regulations?

#### Interaction of discrete with continuous variable

Now let's examine whether the effect of lot size on home value differs depending on the number of rooms. We'll create an interaction between the discrete variable `n_rooms` and the continuous variable `lot_size_acres`.

```{r}
# Model with interaction between n_rooms and lot_size_acres
model_interaction2 <- feols(
  total_value ~ poly(lot_size_acres, 4) +
    i(n_rooms) +
    i(n_stories) +
    i(n_stories, lot_size_acres) +
    building_area,
  data = df,
  vcov = "HC1"
)
summary(model_interaction2)
```

**Questions to consider:**
1. How do you interpret the coefficient on the interaction term `n_rooms::2:lot_size_acres`?
2. Is this interaction statistically significant?
3. What does this tell us about how the value of additional lot size depends on the number of rooms?

*Answer:*



### Fixed Effects

One concern we might have with interpreting our regression results is that the number of rooms a house has is likely related to what town the home is in. 
Let's add a set of town indicators to limit our regression to compare homes "in the same town", so to speak.
It turns out there are 98 different towns in our sample, so this will end up slowing down our regression quite a bit. 
Instead, we will specify the `town` variable after a vertical bar, `|`, in our formula. 
It will look like `y ~ x | town`. 
`feols` will estimate these effects very quickly, but note their coefficients are not displayed in the summary. 

Modify the previous regression model to add town "fixed effects" (i.e. a set of indicators for each town).

```{r}
# Model with town fixed effects
model_fe <- feols(
  total_value ~ poly(lot_size_acres, 4) + i(n_rooms) + building_area | town,
  data = df,
  vcov = "HC1"
)
summary(model_fe)
```

How does this result change our estimate on `building_area`?

*Answer:*



### Zoning Regulation

Zoning regulations limit the kind of development that can occur on a parcel.
We have a bunch of zoning measures, but the one of interest is: 
- `z_mls_acres`, the minimum lot size (the required amount of land needed to build a house)

Using regression methods, try to present a case that more stringent regulations (higher minimum lot size) is associated with larger observed lot sizes. 
Control for `dist_to_cbd` flexibly when doing so since lot size and distance to cbd are correlated.
You should write up your results in a paragraph and include a figure. 
Make sure to interpret the statistical significance of your results.

```{r}
# Regression of lot size on zoning minimum lot size
model_zoning <- feols(
  lot_size_acres ~ z_mls_acres,
  data = df,
  vcov = "HC1"
)
summary(model_zoning)

# Create prediction grid for the regression line
zoning_pred <- data.frame(
  z_mls_acres = seq(
    min(sample_df$z_mls_acres),
    max(sample_df$z_mls_acres),
    length.out = 100
  )
)
zoning_pred$predicted_lot_size <- predict(model_zoning, newdata = zoning_pred)

# Scatterplot of lot size vs zoning minimum using ggplot
ggplot() +
  geom_point(
    aes(x = z_mls_acres, y = lot_size_acres),
    data = sample_df,
    shape = 19,
    alpha = 0.1
  ) +
  geom_line(
    aes(x = z_mls_acres, y = predicted_lot_size),
    data = zoning_pred,
    color = "red"
  ) +
  labs(
    title = "Scatterplot of Lot Size vs Minimum Lot Size Requirement",
    x = "Minimum Lot Size Requirement (Acres)",
    y = "Actual Lot Size (Acres)"
  ) +
  theme_minimal()
```

*Answer:*
