---
title: "PS01 - Statistics Review"
author: "YOUR NAME"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: pdf
---

```{r}
library(tidyverse)
```

# Simulation and Sampling Distributions

In this problem set, we'll explore the central limit theorem and sampling distributions through simulation.

We'll create a general simulation function that can work with different sampling distributions and statistics.

## Creating a Simulation Function

Let's write a function that simulates drawing samples and calculating statistics.

The function will take three main parameters:
- `N`: sample size
- `draw_sample`: a function that draws a sample of size N
- `calculate_statistic`: a function that calculates a statistic from the sample

```{r}
## Helper function that will draw B samples using `draw_sample` and calcuating the statistic for each sample
## Returns the set of statistics generated
simulate_sampling_distribution <- function(
  N,
  draw_sample,
  calculate_statistic,
  B = 2500
) {
  # Create a vector to store the statistics
  statistics <- numeric(B)

  # Run B simulations
  for (i in 1:B) {
    # Draw a sample of size N
    x <- draw_sample(N)

    # Calculate the statistic and store it
    statistics[i] <- calculate_statistic(x)
  }

  return(statistics)
}
```

## Example 1: Normal Distribution

Let's test our function with samples drawn from a normal distribution.
We'll calculate the sample mean.

```{r}
## Sample from normal function
draw_normal_sample <- function(N) {
  rnorm(N, mean = 10, sd = 2)
}

## Calculate the sample mean
calculate_sample_mean <- function(x) {
  mean(x)
}
calculate_sample_mean_using_half_data <- function(x) {
  mean(x[1:(length(x) / 2)])
}
calculate_sample_median <- function(x) {
  median(x)
}


## Run simulation and analyze results
set.seed(123)
sample_means <- simulate_sampling_distribution(
  N = 30,
  draw_sample = draw_normal_sample,
  calculate_statistic = calculate_sample_mean,
  B = 2500
)
sample_medians <- simulate_sampling_distribution(
  N = 30,
  draw_sample = draw_normal_sample,
  calculate_statistic = calculate_sample_median,
  B = 2500
)

ggplot() +
  geom_density(
    aes(x = sample_means, y = after_stat(density), color = "Sample Means"),
    # color = "lightblue",
    bins = 100,
    linewidth = 2
  ) +
  geom_density(
    aes(x = sample_medians, y = after_stat(density), color = "Sample Medians"),
    # color = "lightcoral",
    bins = 100,
    linewidth = 2
  ) +
  labs(
    title = "Sampling Distribution of Sample Mean\n(Normal Population)",
    x = "Sample Mean",
    y = "Density"
  ) +
  theme_minimal()


```

```{r}
## Plot the sampling distribution
ggplot() +
  ## empirical sampling distribution
  geom_histogram(
    aes(x = sample_means, y = after_stat(density)),
    fill = "lightblue",
    color = "black",
    bins = 100
  ) +
  ## theoretical sampling distribution
  stat_function(
    fun = function(x) dnorm(x, mean = 10, sd = 2 / sqrt(30)),
    color = "darkblue",
    linewidth = 1
  ) +
  labs(
    title = "Sampling Distribution of Sample Mean\n(Normal Population)",
    x = "Sample Mean",
    y = "Density"
  ) +
  theme_minimal()
```



## Example 2: Binomial Distribution

Now let's try with a binomial distribution. 
We'll calculate the sample proportion of successes.

```{r}
N <- 30
p <- 0.3
draw_binomial_sample <- function(N) {
  # Bernoulli trials with probability of success = 0.3
  rbinom(N, size = 1, prob = p)
}
# x <- draw_binomial_sample(10000)

calculate_sample_proportion <- function(sample_data) {
  mean(sample_data) # For 0/1 data, mean = proportion
}

## Run simulation and analyze results
set.seed(456)
binomial_props <- simulate_sampling_distribution(
  N = N,
  draw_sample = draw_binomial_sample,
  calculate_statistic = calculate_sample_proportion,
  B = 2500
)

ggplot() +
  ## empirical sampling distribution
  geom_histogram(
    aes(x = binomial_props, y = after_stat(density)),
    fill = "lightblue",
    color = "black",
    bins = 100
  ) +
  ## theoretical sampling distribution
  stat_function(
    fun = function(x) dnorm(x, mean = p, sd = sqrt(p * (1 - p) / N)),
    color = "darkblue",
    linewidth = 1
  ) +
  labs(
    title = "Sampling Distribution of Sample Proportion\n(Binomial Population)",
    x = "Sample Proportion",
    y = "Density"
  ) +
  theme_minimal()
```

#### Exercise

1. What happens when `N` is small? How does the normal approximation of the sampling distribution perform? 



2. Does the number of simulations (B) change the sample distribution? 









## Example 3: Exponential Distribution

Let's try with an exponential distribution, which is skewed.
We'll see how the sample mean behaves.

```{r}
## Define function for exponential distribution
draw_exponential_sample <- function(N) {
  rexp(N, rate = 1) # Mean = 1, variance = 1
}

## Example to see the distribution
hist(draw_exponential_sample(10000), breaks = 100)
```

```{r}
N = 5

## Run simulation and analyze results
set.seed(789)
exp_means <- simulate_sampling_distribution(
  N = N,
  draw_sample = draw_exponential_sample,
  calculate_statistic = calculate_sample_mean,
  B = 2500
)

# Summary of the sampling distribution
summary(exp_means)

## Plot the sampling distribution
ggplot() +
  ## empirical sampling distribution
  geom_histogram(
    aes(x = exp_means, y = after_stat(density)),
    fill = "lightblue",
    color = "black",
    bins = 100
  ) +
  ## theoretical sampling distribution
  stat_function(
    fun = dnorm,
    args = list(mean = 1, sd = 1 / sqrt(N)),
    color = "darkblue",
    linewidth = 1
  ) +
  labs(
    title = "Sampling Distribution of Sample Mean\n(Exponential Population)",
    x = "Sample Mean",
    y = "Density"
  ) +
  theme_minimal()
```

#### Exercise

1.  Try a few different values of `N`, e.g. try 5, 10, 20, 40, and 100. How "quickly" does the normal approximation start to work well?

2. When `N` is small,  the normal approximation is a poor approximation of the true sample distribution of the statistic.
   Why is it particularly problematic when conducting inference using the normal approximation (1.96 * standard error)?
   (Hint: consider the tails / t-distribution) 





## t-statistic

One of the most common statistic we will calculate is the $t$ test-statistic. 
This is because we will often want to test whether a sample mean is statistically significantly different from a hypothesized value.
Let's look at the sample distribution of our test statistic.

```{r}
N = 50

## Define t-statistic function
calculate_t_statistic <- function(x) {
  n <- length(x)
  sample_mean <- mean(x)
  sample_sd <- sd(x)

  # t-statistic: (x̄ - μ) / (s/√n)
  # Here we test against true mean = 10
  (sample_mean - 1) / (sample_sd / sqrt(n))

  ## (sample_sd / sqrt(n)) is called the "standard error"
  ## which is the estimate of the standard deviation of the sampling distribution
}

## Run t-distribution simulation
set.seed(20260120)
t_stats <- simulate_sampling_distribution(
  N = N, # Small sample size
  draw_sample = draw_exponential_sample,
  calculate_statistic = calculate_t_statistic,
  B = 250000
)

## Plot t-statistics
ggplot() +
  ## empirical sampling distribution
  geom_histogram(
    aes(x = t_stats, y = after_stat(density)),
    fill = "lightblue",
    color = "black",
    bins = 100
  ) +
  ## theoretical sampling distribution
  stat_function(
    fun = function(x) dt(x, df = N - 1),
    color = "darkblue",
    linewidth = 1
  ) +
  labs(
    title = "Distribution of t-statistic\n(Small Sample from Normal Population)",
    x = "t-statistic",
    y = "Density",
  ) +
  theme_minimal()
```

Perhaps unsurprisingly, our statistic follows the `t` distribution with $N - 1$ degrees of freedom.



## Exercises

### Exercise 1

Modify the simulation to use a uniform distribution U(0, 10).
Draw $N = 100$
Calculate the sample mean.

```{r}

```

Compare the sampling distribution to the theoretical normal distribution.



### Exercise 2

Create a function to calculate the sample variance (use the `var` function).
Run simulations with different sample sizes ($N = 10$, $N = 30$, $N = 100$).


```{r}

```

How does the sampling distribution of the variance change with sample size?


### Exercise 3

Simulate the sampling distribution of the median for samples from a normal distribution.
Compare it to the sampling distribution of the mean.

```{r}

```

Which sample distribution has the larger variance (use the `var` function)?



## Key Takeaways

For many statistics, the central limit theorem (CLT) will ensure that the sampling distribution is *approximately* normally distributed *in large samples*.
However, this is not guaranteed and caution should be had.
For most distributions, sample sizes of 30 or more are sufficient for the CLT to provide a good approximation, though this varies with the degree of skewness in the population.



## Confidence Intervals Coverage

```{r}
set.seed(123)
does_confidence_interval_contain_true_value <- function(x) {
  (10 >= mean(x) - 1.96 * sd(x) / sqrt(length(x))) &&
    (10 <= mean(x) + 1.96 * sd(x) / sqrt(length(x)))
}
res <- simulate_sampling_distribution(
  N = 100,
  draw_sample = draw_normal_sample,
  calculate_statistic = does_confidence_interval_contain_true_value,
  B = 25000
)

table(res)
mean(res) * 100
```
