---
title: "PS03 - Overfitting"
author: "YOUR NAME"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: pdf
---

```{r}
library(tidyverse)
library(fixest)
```



# Model Specification and Prediction Error

In this problem set, we'll explore how different model specifications perform when the true functional form is unknown and non-linear. 
We'll generate data from a complex data generating process and evaluate how well different regression models capture the underlying relationship.

```{r}
source(
  "https://gist.githubusercontent.com/kylebutts/06908e23880457e2d7fb69444f7c21c0/raw/89003252ff3d248b485e5c7d826dc785b298f0da/dgp.R"
)
```

The above code cell loads our secret data-generating function, which you can call with `draw_sample(N = 500)`.
It returns a dataset with four variables:
- The outcome `y`
- Two continuous covariates, `x1` and `x2`
- A discrete variable called `group`.

The outcome is generated through a non-linear combination of these covariates

## Model Evaluation Function

We'll create a function that generates data, fits a regression model with a specified formula, and returns the model fit measured by Mean Squared Prediction Error (MSPE).

```{r}
## Function to evaluate model specification
evaluate_model <- function(
  formula,
  N = 500,
  B = 100,
  in_sample = TRUE,
  seed = NULL
) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  mspes <- vector("numeric", B)
  for (b in 1:B) {
    ## Generate training data
    train_data <- draw_sample(N = N)

    ## Fit model on training data
    model <- feols(formula, data = train_data)

    if (in_sample) {
      ## In-sample MSPE: predict on training data
      predictions <- fitted(model)
      actuals <- train_data$y

      ## Calculate in-sample MSPE
      mspe <- mean((actuals - predictions)^2)
    } else {
      ## Generate fresh test data (different observations, same size)
      test_data <- draw_sample(N = N)

      ## Predict on test data
      predictions <- predict(model, newdata = test_data)
      actuals <- test_data$y

      ## Calculate out-of-sample MSPE
      mspe <- mean((actuals - predictions)^2)
    }

    mspes[b] <- mspe
  }

  return(mspes)
}
```

## Example 0: Intercept only

To get a sense of how a truly bad model would perform, let's use just an intercept as our formula:

```{r}
baseline_mspe <- evaluate_model(
  formula = y ~ 1,
  N = 500,
  in_sample = TRUE
)

mean(baseline_mspe)
```

## Example 1: Simple Linear Model


Let's start with a simple linear specification and evaluate its performance.

```{r}
simple_linear_mspe <- evaluate_model(
  formula = y ~ x1 + x2 + i(group),
  N = 500,
  in_sample = TRUE
)

mean(simple_linear_mspe)
```

The simple linear model provides a baseline. However, since the true relationship is non-linear, we expect this specification to miss important patterns in the data.

## Example 2: Adding Polynomial Terms

Let's improve our model by adding quadratic terms to capture curvature in the relationship.

```{r}
poly_mspe <- evaluate_model(
  formula = y ~ x1 + I(x1^2) + x2 + I(x2^2) + i(group),
  N = 500,
  in_sample = TRUE
)

mean(poly_mspe)
```

By adding polynomial terms, we allow the model to capture non-linear relationships between the continuous covariates and the outcome. Compare the MSPE to the simple linear model.


---

# In-Class Challenge: Find the Best Model

**Challenge**: Using the `evaluate_model` function, find the regression model with the **lowest MSPE**. The student with the best model will win a prize!

**Rules**:
- You can use any combination of `x1`, `x2`, and `i(group)`
- You can include polynomial terms (e.g., `I(x1^2)`, `I(x1^3)`)
- You can include interaction terms (e.g., `x1 * x2`, `i(group, x1)`)
- Use `N = 500` and `set.seed(20260210)` for consistency
- The goal is to minimize the MSPE

**Tips**:
- Try different combinations of main effects
- Experiment with polynomial degrees
- Test various interaction structures
- Consider how `group` might modify the effects of `x1` and `x2`

## Your Best Model

Once you've found your best model, report it below. Make sure to use `set.seed(2025)` and `N = 500`!
Use the tips below for adding different kinds of terms in your model

```{r}
## Experiment with different model specifications here
## Try different combinations of variables, polynomials, and interactions
my_best_formula <- y ~ x1 + x2 + i(group) ## Change this!

my_best_mspe <- evaluate_model(
  formula = my_best_formula,
  N = 500,
  seed = 20260210
)

cat("My best in-sample MSPE:", mean(my_best_mspe), "\n")
cat("Formula:", deparse(test_formula), "\n")
```


## Tips: Flexible Model Specification with `fixest`

The `fixest` package provides powerful tools for flexible model specification. Here are some useful techniques:

### Polynomial Terms

Use `I()` to include polynomial terms:

```{r, eval=FALSE}
## Quadratic specification
formula = y ~ x1 + I(x1^2) + x2 + I(x2^2) + i(group)

## Cubic specification
formula = y ~ x1 + I(x1^2) + I(x1^3) + x2 + i(group)
```

### Interactions Between Continuous Variables

Use `:` to interact two continuous variables without including the main effects automatically:

```{r, eval=FALSE}
## Interaction only (no main effects)
formula = y ~ x1:x2 + i(group)

## Interaction with main effects (equivalent to x1 * x2)
formula = y ~ x1 + x2 + x1:x2 + i(group)
```

### Categorical Indicators

Use `i()` to create indicator variables for categorical covariates:

```{r, eval=FALSE}
## Create indicators for group categories (omits first level by default)
formula = y ~ x1 + x2 + i(group)

## Same as: y ~ x1 + x2 + i(group) (since group is already categorical)
```

### Varying Slopes (Interactions with Categories)

Use `i(cat, var)` to allow the slope of a continuous variable to vary by category:

```{r, eval=FALSE}
## Allow x1 slope to differ by group category
formula = y ~ x1 + x2 + i(group, x1)

## This creates: x1 + x2 + i(group) + x1:group_B + x1:group_C
## (slope of x1 differs for each level of group)

## Combine with varying intercepts and slopes
formula = y ~ x2 + i(group) + i(group, x1)
## Creates: x2 + i(group) + x1:group_B + x1:group_C
```


















--- 






















## Out-of-Sample Performance

Now, let's see how well your model actually performs. 
Evaluate your **best model** (the one you submitted above) on fresh testing data by setting `in_sample = FALSE`.

This will train your model on one set of 500 observations and evaluate it on a completely different set of 500 observations.

```{r}
## Evaluate YOUR best model on out-of-sample data
my_best_in_sample <- evaluate_model(
  formula = my_best_formula,
  N = 200,
  in_sample = TRUE,
  seed = 20260204
)

my_best_out_sample <- evaluate_model(
  formula = my_best_formula,
  N = 200,
  in_sample = FALSE,
  seed = 20260204
)

cat(
  "My best model in-sample MSPE:",
  round(mean(my_best_in_sample), 3),
  "\n"
)
cat(
  "My best model out-of-sample MSPE:",
  round(mean(my_best_out_sample), 3),
  "\n"
)
```

How did your model perform? Did the out-of-sample MSPE surprise you?



#### Exercise

1. Which model won the in-sample competition? How complicated was the winning formula?



2. How did the winning model perform out-of-sample? Did it generalize well, or did it overfit?










# Understanding Overfitting

## Bias-Variance Tradeoff at Fixed Sample Size

So far, we've compared specific models. Now let's systematically explore how model complexity affects prediction error for a fixed sample size. As we increase model complexity (adding higher-order polynomials and more interactions), we expect to see:

- **In-sample MSPE**: Should continuously decrease or stay flat (more flexible models can always fit training data at least as well)
- **Out-of-sample MSPE**: Should initially decrease (reducing bias) but eventually increase (overfitting, increasing variance)

This U-shaped pattern for out-of-sample error illustrates the **bias-variance tradeoff**.

```{r}
## Function to evaluate models with increasing polynomial complexity
evaluate_complexity <- function(
  N = 500,
  max_degree = 6,
  B = 100,
  seed = NULL
) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  results <- data.frame(
    degree = integer(),
    model_type = character(),
    mspe_type = character(),
    mspe = numeric()
  )

  for (degree in 1:max_degree) {
    ## Build formula with polynomial terms up to current degree
    ## Include interactions between x1 and x2 polynomials
    poly_terms_x1 <- paste0("I(x1^", 1:degree, ")", collapse = " + ")
    poly_terms_x2 <- paste0("I(x2^", 1:degree, ")", collapse = " + ")

    ## Add interactions between all polynomial terms
    if (degree >= 2) {
      interaction_terms <- c()
      for (i in 1:degree) {
        for (j in 1:degree) {
          if (i + j <= degree + 1) {
            ## Cap interaction order
            interaction_terms <- c(
              interaction_terms,
              paste0("I(x1^", i, ")", ":", "I(x2^", j, ")")
            )
          }
        }
      }
      interaction_str <- paste(interaction_terms, collapse = " + ")
      formula_str <- paste(
        "y ~",
        poly_terms_x1,
        "+",
        poly_terms_x2,
        "+",
        interaction_str,
        "+ i(group)"
      )
    } else {
      formula_str <- paste(
        "y ~",
        poly_terms_x1,
        "+",
        poly_terms_x2,
        "+ i(group)"
      )
    }

    formula <- as.formula(formula_str)

    ## Evaluate in-sample
    mspe_in <- evaluate_model(formula, N = N, B = B, in_sample = TRUE)

    ## Evaluate out-of-sample
    mspe_out <- evaluate_model(formula, N = N, B = B, in_sample = FALSE)

    ## Count parameters (approximate)
    n_params <- degree * 2 ## x1 and x2 polynomials
    if (degree >= 2) {
      n_params <- n_params + length(interaction_terms)
    }
    n_params <- n_params + 2 ## group indicators

    results <- rbind(
      results,
      data.frame(
        degree = rep(degree, 2),
        n_params = rep(n_params, 2),
        mspe_type = c("In-sample", "Out-of-sample"),
        mspe = c(mean(mspe_in), mean(mspe_out))
      )
    )
  }

  return(results)
}

## Run the complexity evaluation
N = 30
complexity_results <- evaluate_complexity(
  N = N,
  max_degree = 6,
  B = 50,
  seed = 20260210
)

## Plot the bias-variance tradeoff
ggplot(complexity_results, aes(x = degree, y = mspe, color = mspe_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Bias-Variance Tradeoff",
    subtitle = paste("Fixed sample size N = ", N),
    x = "Polynomial Degree (Model Complexity)",
    y = "Mean Squared Prediction Error",
    color = "Evaluation Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

#### Exercise

1. **Interpreting the Bias-Variance Tradeoff Plot**: 
  - At what degree does the out-of-sample MSPE reach its minimum? 
  
  - Why does the in-sample MSPE continue to decrease (or flatten) as complexity increases?
  
  - What happens to the gap between in-sample and out-of-sample MSPE as model complexity increases?



2. **Optimal Complexity**: The true data generating process uses degree-2 polynomials for $x_1$ and $x_2$. Does the minimum out-of-sample MSPE occur at degree 2? If not, why might a more complex model show better out-of-sample performance?




## Sample Size and Model Complexity

The bias-variance tradeoff depends on sample size. With more data, we can support more complex models without overfitting.

```{r}
## Function to compare models across sample sizes
compare_models_by_N <- function(
  N_values,
  formula_simple,
  formula_complex
) {
  results <- data.frame(
    N = numeric(),
    model = character(),
    mspe_type = character(),
    mspe = numeric()
  )

  for (N in N_values) {
    ## Simple model
    simple_in <- evaluate_model(
      formula_simple,
      N = N,
      in_sample = TRUE
    )
    simple_out <- evaluate_model(
      formula_simple,
      N = N,
      in_sample = FALSE
    )

    ## Complex model
    complex_in <- evaluate_model(
      formula_complex,
      N = N,
      in_sample = TRUE
    )
    complex_out <- evaluate_model(
      formula_complex,
      N = N,
      in_sample = FALSE
    )

    results <- rbind(
      results,
      data.frame(
        N = rep(N, 4),
        model = c("Simple", "Simple", "Complex", "Complex"),
        mspe_type = c(
          "In-sample",
          "Out-of-sample",
          "In-sample",
          "Out-of-sample"
        ),
        mspe = c(
          mean(simple_in),
          mean(simple_out),
          mean(complex_in),
          mean(complex_out)
        )
      )
    )
  }

  return(results)
}

## Run comparison (using small B for speed)
N_values <- c(100, 200, 500, 1000, 2500, 5000)
set.seed(20260210)
comparison_results <- compare_models_by_N(
  N_values = N_values,
  formula_simple = y ~ x1 + x2 + i(group),
  formula_complex = y ~ i(group) +
    i(group, x1) +
    i(group, x1^2) +
    i(group, x1^3) +
    i(group, x1^4) +
    i(group, x2) +
    i(group, x2^2) +
    i(group, x2^3) +
    i(group, x2^4)
)

## Plot results
ggplot(
  comparison_results |> filter(N >= 200),
  aes(x = factor(N), y = mspe, color = model, linetype = mspe_type)
) +
  geom_line(aes(group = interaction(model, mspe_type)), linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Model Performance by Sample Size",
    subtitle = "Comparing Simple vs Complex Specifications",
    x = "Sample Size (N)",
    y = "Mean MSPE",
    color = "Model",
    linetype = "MSPE Type"
  ) +
  theme_minimal()
```


#### Exercise

1. How does the gap between in-sample and out-of-sample performance change as sample size increases for the complex model? What does this tell us about the bias-variance tradeoff?












## Key Takeaways

1. **Model Specification Matters**: When the true functional form is unknown, different specifications can lead to very different prediction performance.

2. **In-Sample vs Out-of-Sample**: In-sample fit can be misleading due to overfitting. Out-of-sample evaluation provides a more honest assessment of how well the model will perform on new data.

3. **Bias-Variance Tradeoff**: Simple models may underfit (high bias), while complex models may overfit (high variance). The optimal model balances these concerns.

4. **Sample Size**: With larger samples, we can afford more complex models without overfitting. With small samples, simpler models often generalize better.



